DynaProg for Scala

### 2013-01-15 docs/report/design.tex
Reports: SWat problems.

ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase). Such separation makes possible to define multiple algebra for the same grammar. This has two main applications:\ol
\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Watermann share the same grammar but have a different evaluation algebra
\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data. For example in matrix chain multiplication, the first step solves the underlying dynamic problem by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
\ole

>>>>

ADP is a formalization of parsers that introduces a distinction between the \textbf{parsing grammar} (recognition phase) and an associated \textbf{algebra} (evaluation phase). Such separation makes possible to define multiple algebra for the same grammar. This has two main applications:\ol
\item Experiment variants with the same grammar:  for example, Needleman-Wunsch and Smith-Waterman share the same grammar but have a different evaluation algebra
\item Use an evaluation and execution algebra: a dynamic programming problem is solved in two steps: computing one optimal solution and applying it over actual data. For example in matrix chain multiplication, the first step solves the underlying dynamic problem by evaluating the number of necessary multiplications, the second step \textit{effectively} multiplies matrices according to the order previously defined.
\ole

### 2013-01-15 docs/report/problems.tex
Reports: SWat problems.

\subsubsection{Definitions}\ul
\item \textbf{Cost or score:} refers to the result of the dynamic programming recurrence formula.
\item \textbf{Backtrack:} the backtrack is the information related to a score that describe how it has been obtained by referring to immediately previous elements. By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
\item \textbf{Alphabets:} the alphabet represent an enumeration of the possible values. Their size helps determining how many bits are required in the implementation to store the values. Alphabets must be defined for input, wavefront, cost and backtrack.
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer indifferently by the matrix or the matrices to all the intermediate cost- and backtrack-related informations that are necessary to solve the dynamic programming problem of interest. Matrices elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and or backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram, possibly cropped at its parent matrix boundaries.
\item \textbf{Wavefront:} the wavefront consists of all the data necessary to reconstruct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i-1,j)}, M_{(i-2,j-1)}\big)$ has delay 3).
\ule

\subsubsection{Litterature classification}
In \cite{gpu_atlp}, dynamic programming problems are classified according to two criteria:\ul
\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman). When two or more terms appear, the problem is polyadic (ex: Fibonacci, $F_n = F_{n-1} + F_{n-2}$).
When a problem is polyadic with index $p$, it also means that its backtracking forms a $p$-ary tree (where each node has at most $p$ children).

\item \textbf{Serial/non-serial:} a problem is serial ($s=0$) when the solutions depends on a fixed number of previous solutions (ex: Fibonacci), otherwise it is said to be non-serial ($s\ge 1$), as the number of dependencies grows with the size of the subproblem. That is computing an element of the matrix would require $O(n^s)$.  (ex: Smith-Waterman with arbitrary gap is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula)
	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
\ule

Note that the algorithmic complexity of a problem is exactly $O\big(n^{d+s}\big)$.

\subsubsection{Calculus simplifications} \label{calc_simplifications}
In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]

@@

% ------------------------------------------------------------------------------------------------
\subsection{Problems of interest}
We usually focus on problem that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$). Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability. Also we tend to limit algorithmic complexity of the problems as from $O(n^4)$ on, running time becomes a severely limiting factor.

@@

Although not necessary to understand the project, the description of some of the most common dynamic programming problem is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems. Would the reader be familiar with dynamic programming, he could immediately jump to the next section.

A tighter analysis on the alphabet and intermediate results size is done because FPGA was also considered as a possible execution platform.\footnote{Initially, the project envisioned a collaboration with \href{mailto:nithin.george@epfl.ch}{Nithin George}, a PhD student at Processor Laboratory (LAP), EPFL. Unfortunately, at the time of writing, progress in this direction is not sufficient to be covered in this report. However, FPGA is considered as a future deployment platform for a particular class of problems.}

@@

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman (simple)}\label{swat_simple}\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		0 & stop\\
		M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
		M_{(i-1,j)}-d & N\\
		M_{(i,j-1)}-d & W
	\end{array}\right\}=B_{(i,j)} \]

@@

\item Optimizations:\ul
	\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built). This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
	\item Since first line and column of the matrix (corresponding to a matching with one string being empty) are zeroed, their initialization might be omitted, but this would implies more involved initialization and computations, which is cumbersome.
	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$, we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned. Hence valid input indices would become $S[1 \cdots m]$ and $T[1 \cdots n]$.
	\ule
\ole

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with affine gap extension cost} \label{swat_affine}
\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.
\item Matrices: $M_{(m+1) \times (n+1)}, E_{(m+1) \times (n+1)}, F_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

@@

Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules whose kernel is
\[ [M;E;F]_{(i,j)} = f_{\rm kernel} ( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} ) \]

\item Backtracking: similar to \nameref{swat_simple}
\item Visualization: same as \nameref{swat_simple}
\item Optimizations: Notice that this recurrence is very similar to \nameref{swat_simple} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$). Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of two additional matrices.
\ole

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman with arbitrary gap cost}\label{swat_arbitrary}\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap. Without loss of generality, let $m\ge n$ (swap inputs if necessary). Example penalty function could be\footnote{Intuition: long gaps penalize less, at some point, one large gap is better than matching with smaller gaps.} $g(x)=m-x$.
\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
	\ule

@@

\item Backtracking: similar to \nameref{swat_simple} except that you can jump of $k$ cells along the rows or along the columns.
\item Visualization:
\CFig{Smith-Waterman (arbitrary gap cost) dependencies}{
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
	\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,-1]{5,5}{0.8}
}

@@

The recurrence consists of two non-serial dependencies as in \textit{\nameref{swat_arbitrary}} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{figure}[H]\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}

@@

The research on ad-hoc implementation has focused on three kind of problems:\ul
\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
\item Biological sequence alignment (Smith-Watermann) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
\ule

>>>>

% ----------------------------------------------
\subsubsection{Definitions}\ul
\item \textbf{Cost or score:} refers to the result of the dynamic programming recurrence formula.
\item \textbf{Backtrack:} the backtrack is the information related to a score that describe how it has been obtained by referring to immediately previous elements. By induction on the backtrack, the \textbf{trace} (that describe \textit{all} thee steps to obtain the result) can be obtained.
\item \textbf{Alphabets:} an alphabet is an enumeration of the possible values. Its size helps determining how many bits are required in the implementation to store all values. Alphabets are defined for input, cost, backtrack and wavefront.
\item \textbf{Dimensions:} let $n$ the size of the input and $d$ the dimension of the underlying matrix.
\item \textbf{Matrices:} we refer by \textit{matrix} or \textit{matrices} to all the memoized intermediate cost- and backtrack-related information that is necessary to solve the dynamic programming problem of interest. Matrix elements are usually denoted by $M_{(i,j)}$ ($i^{\rm th}$ line , $j^{\rm th}$ column).
\item \textbf{Computation block:} this is a part of the DP matrix (cost and backtrack) that we want to compute. A block might be either a sub-matrix (rectangular) or a parallelogram (possibly reduced by taking the intersection with its enclosing matrix).
\item \textbf{Wavefront:} the wavefront consists of the minimum data necessary to construct a computation block of the DP matrix. It might include some previous lines/columns/diagonals as well as line-/column-/diagonal-wise aggregations (min, max, sum, ...).
\item \textbf{Delay:} we call delay the maximum distance between an element and its dependencies along column and lines (ex: recurrence $M_{(i,j)}=f\big(M_{(i+1,j)}, M_{(i+2,j-1)}\big)$ has delay 3).
\ule

% ----------------------------------------------
\subsubsection{Litterature classification}
In \cite{gpu_atlp}, dynamic programming problems are classified according to two criteria:\ul
\item \textbf{Monadic/polyadic:} a problem is monadic when only one of the previously computed term appears in the right hand-side of the recurrence formula (ex: Smith-Waterman \S\ref{swat_simple}). When two or more terms appear, the problem is polyadic (ex: Fibonacci, $F_n = F_{n-1} + F_{n-2}$).
When a problem is polyadic with index $p$, it also means that its backtracking forms a $p$-ary tree (where each node has at most $p$ children).

\item \textbf{Serial/non-serial:} a problem is serial ($s=0$) when the solutions depends on a fixed number of previous solutions (ex: Fibonacci), otherwise it is said to be non-serial ($s\ge 1$), as the number of dependencies grows with the size of the subproblem. That is computing an element of the matrix would require $O(n^s)$. For example, Smith-Waterman with arbitrary gap cost (\S\ref{swat_arbitrary}) is $s=1$; we can usually infer $s$ from the number of bound variables in the recurrence formula (see recurrence formulae in \S\ref{problem_examples}).
%	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
\ule
Note that the algorithmic complexity of a problem is exactly $O\big(n^{d+s}\big)$.

% ----------------------------------------------
\subsubsection{Recurrence formulae simplifications} \label{calc_simplifications}
In some special cases, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]

@@

% ------------------------------------------------------------------------------------------------
\subsection{Problems of interest} \label{problem_examples}
We here focus on problems that have an underlying bi-dimensional matrix ($d=2$) because they can be parallelized (as opposed to be serial if $d=1$) and can solve large problems (of size $n$). Problems of higher matrix dimensionality ($d\ge3$) require substantial memory which severely impacts their scalability. Also it seems that most problems of interest have an algorithmic complexity of at most $O(n^4)$, probably because running time would otherwise becomes a severely limiting factor for the size of the problem.

@@

Although not necessary to understand the project, the description of some of the most common dynamic programming problems is relevant to capture the essence of the dynamic programming processes and be able to compare and search for similarities among problems. Would the reader be familiar with dynamic programming, he could immediately jump to the next section.

A tighter analysis on the alphabet and intermediate results size is done because FPGA was also considered as a possible execution platform.

@@

% ----------------------------------------------
\newpage
\subsubsection{Smith-Waterman (simple)}\label{swat_simple}
Smith-Waterman is a biological sequence alignment algorithm. It tries to find the maximum number of correspondences between two DNA sequences; variants of this algorithm include \href{http://en.wikipedia.org/wiki/Needleman–Wunsch_algorithm}{Needleman-Wunsch}, and minimum edit distance family that generalizes on strings (\href{http://en.wikipedia.org/wiki/Hamming_distance}{Hamming distance}, \href{http://en.wikipedia.org/wiki/Levenshtein_distance}{Levenshtein distance}, ...). We explore three variants of this algorithm: simple (\S\ref{swat_simple}), affine (\S\ref{swat_affine}) and arbitrary (\S\ref{swat_arbitrary}) gap cost models. We study this problem because it has the interesting properties of using multiple input sequences and being suitable for hardware generation \cite{swat_fpga}.\\[6pt]\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$, with constant mismatch penalty ($d$) and arbitrary matching function (${\rm cost}(\_,\_)$).
\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence: \[M_{(i,j)}=\max\left\{\begin{array}{l|l}
		0 & stop\\
		M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
		M_{(i-1,j)}-d & N\\
		M_{(i,j-1)}-d & W
	\end{array}\right\}=B_{(i,j)} \]

@@

\item Optimizations:\ul
	\item In serial (monadic) problems we can avoid building the matrix $M$ by maintaining only the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being currently built). This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory, where $k$ is the maximum backward lookup.
	\item We could eliminate the first line and column of the matrix as they are filled with zeroes (representing a match with empty string), however this implies more involved computations, which is cumbersome.
	\item Padding: since to fill the $i^{\rm th}$ row we refer to the $(i-1)^{\rm th}$ character of string $S$, we could prepend to both $S$ and $T$ an unused character, so that matrix and input lines are aligned. Hence valid input indices would become $S[1 \cdots m]$ and $T[1 \cdots n]$.
	\ule
\ole

% ----------------------------------------------
\subsubsection{Smith-Waterman with affine gap extension cost} \label{swat_affine}
\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$, where creating a gap in either sequence has an opening penalty ($\alpha$) and an extension penalty ($\beta$).
\item Matrices: $M_{(m+1) \times (n+1)}, E_{(m+1) \times (n+1)}, F_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
	\ule
\item Initialization:\ul
	\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
	\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
	\item S-gap extension cost matrix: $F_{(0,j)}= 0$
	\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
	\ule
\item Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i-1),T(j-1)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

@@

Although the latter notation seems more explicit, it introduces non-serial dependencies that the former set of recurrences is free of. So we need to implement the former rules as
\[ [M;E;F]_{(i,j)} = f \big( [M;E]_{(i,j-1)}, [M;F]_{(i-1,j)}, M_{(i-1,j-1)} \big) \]

\item Backtracking and visualization are similar to \S\ref{swat_simple}
\item Optimizations: Notice that this recurrence is very similar to \S\ref{swat_simple} except that we propagate 3 values ($M,E,F$) instead of a single one ($M$). Also notice that it is possible to propagate $E$ and $F$ inside a respectively horizontal and vertical wavefront, hence removing the need of the two additional matrices.
\ole

% ----------------------------------------------
\subsubsection{Smith-Waterman with arbitrary gap cost}\label{swat_arbitrary}\ol
\item Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$ with an arbitrary gap function $g(x)\ge 0$ where $x$ is the size of the gap. For example\footnote{Intuition: long gaps should penalize less; one large gap might be better than matching with smaller gaps.}: $g(x)=\max(m,n)-x$.

\item Matrices: $M_{(m+1) \times (n+1)}, B_{(m+1) \times (n+1)}$
\item Alphabets:\ul
	\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
	\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_,\_)}) \cdot \min(m,n)$
	\item Backtrack matrix: $\Sigma(B)=\{stop,NW,N_{\{0..m\}},W_{\{0..n\}}\}$
	\ule

@@

\item Backtracking: similar to \S\ref{swat_simple} except that you can jump of $k$ cells along the rows or along the columns.
\item Visualization:
\CFig{Smith-Waterman (arbitrary gap cost) dependencies}{
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd[0,-1]{6,7}{2.8}\Cd[0,-1]{6,6}{1.8}\Cd{6,5}{0.8}
	\Cd[1,0]{0,4}{5.8}\Cd[1,0]{1,4}{4.8}\Cd[1,0]{2,4}{3.8}\Cd[1,0]{3,4}{2.8}\Cd[1,0]{4,4}{1.8}\Cd[1,0]{5,4}{0.8}
	\Cd[1,-1]{5,5}{0.8}
}

@@

The recurrence consists of two non-serial dependencies as in \S\ref{swat_arbitrary} plus a bounded 2-dimensional dependency for bulges.\\[12pt]
Since this problem is non-trivial to understand from the recurrences, we propose an additional illustration of a RNA chain folded according to the Zuker folding algorithm.
\begin{figure}[H]\begin{center}
\includegraphics[width=10cm]{inc/zuker_struct.pdf} \\
\small Types of structural features modeled by the Zuker folding algorithm include: dangling ends (1), \\ internal loop (11), stack (23), multi-loop (47), bulge (68) and hairpin loop (78).\end{center}\caption{An example of an RNA folded into a secondary structure}\end{figure}

@@

The research on ad-hoc implementation has focused on three kind of problems:\ul
\item Genral problems, attempting to provide the most efficient implementation for a particular problem \cite{gpu_atlp}, \cite{robust_mapping}, \cite{gpu_from_dsl}.
\item RNA sequence folding (variants of the Zuker folding): \cite{nussinov_gpu}, \cite{gpu_rnafold}.
\item Biological sequence alignment (Smith-Waterman) for huge sequences: \cite{swat_gpu}, \cite{swat_linear} \cite{swat_mega}.
\ule
